### 2024-08-26 docs/manuscript.tex
Further manuscript edits.

% The abstract could benefit from a clearer statement of the research question, methodology, and key findings. It currently mixes multiple ideas without a clear flow.
\Abstract{
  The development of accurate bioinformatic software tools is crucial
  for the effective analysis of complex biological data. This study
  examines the relationship between the academic department
  affiliations of authors and the accuracy of the bioinformatic tools
  they develop. By analyzing a corpus of previously benchmarked
  bioinformatic software tools, we mapped bioinformatic tools to the
  academic fields of the corresponding authors and evaluated tool
  accuracy by field. Our results suggest that "Medical Informatics,"
  under the broader category of "Technologies," outperforms other
  fields in software accuracy, with a mean proportion of wins in
  accuracy rankings exceeding the null expectation. In contrast, tools
  developed by authors affiliated with "Bioinformatics" and
  "Engineering" fields tend to be less accurate. However, after
  correcting for multiple testing, none of the results are
  statistically significant ($p>0.05$). Our findings reveal no
  association between academic field and bioinformatic
  software accuracy, suggesting that the development of
  interdisciplinary software applications can be effectively
  undertaken by any department with sufficient resources and training.
}

@@

The interdisciplinarity of bioinformatics enables the integration of
methods and perspectives from various disciplines, fostering novel
insights and solutions that might not be possible within a single
field \cite{mazzocchi2019scientific}. This requires a deep
understanding of biological sciences for interpreting
data, and strong computational skills are required to develop
algorithms and analyze the results at scale.

The \textbf{biological and health sciences,} in particular genetics,
biochemistry and molecular biology provide essential domain knowledge,
ensuring that the developed software tools are both biologically relevant
and accurate.  Biologists are well-positioned to identify critical biological
questions and ensure that computational tools are designed to
effectively address these. However, biology departments may
lack the advanced computational expertise required for developing sophisticated
software. In our analysis, we have grouped the biological and health fields
under the category of ``\textbf{domain experts}''.

@@

As the field of bioinformatics evolves and reliance on research
software tools grows with increasing data volumes, this study aims to
determine whether the academic department of a corresponding author
influences the accuracy of the bioinformatic software tools they
develop. Specifically, we examine whether the presumed subject
expertise, reflected in the author's departmental affiliation, has a
measurable impact on tool accuracy, utilizing a benchmarked corpus of
bioinformatic software tools to evaluate this relationship.

@@

We obtained previously benchmarked and collated accuracy rankings from
a published supplement \cite{gardner2024}. Using these rankings, we
mapped the addresses of corresponding authors for each published
software tool to a standardized list of specific ``fields of study''
\cite{fields2014}. These fields were then grouped into higher-level
general fields and broader areas of expertise based on the address
details of each corresponding author. Figure~\ref{fig:fig1} illustrates
the number of tools representing each general and specific field. Most
bioinformatic software is developed by corresponding authors who list
Genetics, Bioinformatics, Computer Science, or similar departments as
their primary address (Figure~\ref{fig:fig1}A). Among the general fields,
the Biological Sciences produce the most software tools, followed by
the Computer Sciences (Figure~\ref{fig:fig1}B).

@@

We tested the assumption that the speciality of academic departments
reflect the quality of the research software they produce. Our
findings showed no significant association, after correcting for
multiple-testing, between the type of academic expertise -- whether
biological domain expert, software developer, or interdisciplinary
scientist -- and the accuracy rankings of bioinformatic software
tools. This leads us to conclude that assumptions linking academic
department affiliation to the quality of bioinformatic software are
likely incorrect. Similarly, both general and specific research fields
showed no significant associations (Figure~\ref{fig:fig2}).

@@

There is often a disconnect between a bioinformatic tool developer's
training and their host department. For example, the author of this
study completed an undergraduate degree in mathematics, a PhD in
bioinformatics, postdoctoral fellowships in computer science and
molecular biology departments, worked for several years at a genomics
research institute, and is currently employed in a Biochemistry
Department. The ``Biochemistry'' departmental label does not accurately
reflect the author's training or recent publications. In fact, the
author admits to not recalling key details of glycolysis, the citric
acid cycle, or the structures of amino acids and proteins.

@@

>>>>

% The abstract could benefit from a clearer statement of the research question, methodology, and key findings. It currently mixes multiple ideas without a clear flow.
\Abstract{ The development of accurate bioinformatic software tools is
  crucial for the effective analysis of complex biological data. This
  study examines the relationship between the academic department
  affiliations of authors and the accuracy of the bioinformatic tools
  they develop. By analyzing a corpus of previously benchmarked
  bioinformatic software tools, we mapped bioinformatic tools to the
  academic fields of the corresponding authors and evaluated tool
  accuracy by field. Our results suggest that ``Medical Informatics''
  outperforms all other fields in bioinformatic software accuracy,
  with a mean proportion of wins in accuracy rankings exceeding the
  null expectation. In contrast, tools developed by authors affiliated
  with "Bioinformatics" and "Engineering" fields tend to be less
  accurate. However, after correcting for multiple testing, none of
  the results are statistically significant ($p>0.05$). Our findings
  reveal no strong association between academic field and
  bioinformatic software accuracy. These findings suggest that the
  development of interdisciplinary software applications can be
  effectively undertaken by any department with sufficient resources
  and training.  }

@@

Bioinformatics integrates methods from various disciplines, fostering
novel insights that are unattainable within a single field
\cite{mazzocchi2019scientific}. This requires a deep understanding of
biological sciences for interpreting data, and strong computational
skills are required to develop algorithms and analyze the results at
scale.

Experts in \textbf{biological and health sciences}, such as genetics
and molecular biology, can ensure that software tools are biologically
relevant and accurate by identifying critical questions and guiding
computational tool development. However, biology departments may lack
the advanced computational expertise required for developing
sophisticated software. In our analysis, we have grouped the
biological and health fields under the category of ``\textbf{domain
  experts}''.

@@

The primary objective of this study is to investigate whether the
academic department affiliation of a corresponding author is
associated with the accuracy of the bioinformatic software tools they
develop. We aim to determine if tools developed by authors from
certain academic fields, such as ``Bioinformatics'', ``Computer
Science'' or ``Genetics'' show higher or lower accuracy compared to
those from other fields. To achieve this, we analyzed a benchmarked
corpus of bioinformatic software tools and evaluated their accuracy
based on the academic affiliations of their developers.

@@

We used a published supplement \cite{gardner2024} to obtain previously
benchmarked accuracy rankings and mapped corresponding authors'
addresses to a standardized list of specific 'fields of study'
\cite{fields2014}. These were further grouped into broader general
fields and areas of expertise. Figure~\ref{fig:fig1} illustrates the
number of tools representing each general and specific field ($N\ge 10$). Most
bioinformatic software is developed by corresponding authors who list
Genetics, Bioinformatics, Computer Science, or similar departments as
their primary address (Figure~\ref{fig:fig1}A). Among the general
fields, the Biological Sciences produce the most software tools,
followed by the Computer Sciences (Figure~\ref{fig:fig1}B).

@@

We tested the assumption that the speciality of academic departments
reflect the quality of the research software they produce.  After
correcting for multiple testing, we found no significant association
between academic expertise type and the accuracy of bioinformatic
tools, suggesting that academic department affiliation does not
correlate with software quality. Similarly, both general and specific
research fields showed no significant associations
(Figure~\ref{fig:fig2}).

@@

There is often a disconnect between a developer’s training and their
departmental affiliation, as evidenced by this author whose diverse
academic and professional background spans mathematics,
bioinformatics, computer science, molecular biology, and genomics, yet
currently lists their affiliation with a Biochemistry Department. The
``Biochemistry'' departmental label does not accurately reflect the
author's training or recent publications. In fact, the author admits
to not recalling key details of glycolysis, the citric acid cycle, or
the structures of amino acids and proteins.

@@

In conclusion, this study does not find strong evidence that the
academic department affiliation of authors is associated with the
accuracy of bioinformatic software tools. This challenges the
assumption that departmental expertise correlates with
software quality. Future research could explore other factors, such
as the nature of interdisciplinary collaborations or the training
of developers, to better understand what factors contribute to
the development of high-quality bioinformatic tools.
